\bf Manifold Clustering for Latent Structure Block Models

### 2022-04-20 draft/draft.Rmd
still working on manuscript

bibliography: bibliography.bib
output: rticles::asa_article
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
- \usepackage{amsfonts}
---

@@

@@

@@

Here, we provide two algorithms for MBM community detection.

First, we consider the case in which communities correspond to manifolds in the latent space that do not intersect and are separated by some finite distance. In this scenario, we use the convergence of the ASE to show that single linkage clustering on the latent space produces a clustering such that the total number of misclustered vertices goes to zero, with high probability.

@@

### Preliminary Theory

#### Distributions of differences of order statistics

Let $D_i = X_{(i+1)} - X_{(i)}$. Then if $\max_i D_i < \delta$, we have sufficient separation of points in $\mathcal{M}_1$. 
Then it is sufficient to quantify $P(\max_i D_i > \delta)$ as a function of $n$ and $\delta$ and show that this converges to zero as $n$ grows to $\infty$.

#### Uniform case

\begin{theorem}
Let $X_1, ..., X_n \iid \Uniform(0, 1)$. 
Then for any $\epsilon$ and $\delta > 0$, 
there exists an $N = O \big(\frac{-\log \epsilon}{\delta} \big)$ such that 
$P(\max_i X_{(i+1)} - X_{(i)} < \delta) \geq 1 - \epsilon$ when $n > N$.
\end{theorem}

#### General case for one-dimensional manifolds

\begin{theorem}
\label{thm:generalized}
Let $X_1, ..., X_n \iid F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
Let $D_i = X_{(i+1)} - X_{(i)}$. 
Then for any $\epsilon > 0$, there exists $N > 0$ such that $P(\max_i D_i < \delta) \geq 1 - \epsilon$ when $n > N$.
\end{theorem}

This converges to $1$ in the limit $n \to \infty$.

We can also approximate $F(1 - \delta) \approx 1 - a \delta$, which yields 
$1 - (n - 1) (1 - a \delta)^n$. Setting this $\geq 1 - \epsilon$ ...
\end{proof}

#### Extension to multidimensional manifolds

\begin{theorem}
\label{thm:multidim}
Let $X_1, ..., X_n \iid F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support. 
Define $E_n$ as the event that an $\eta$-neighborhood graph constructed from the sample is connected. 
Then for any $\epsilon > 0$, there exists $N = O \bigg( \frac{\log \epsilon \eta^r / r^{r/2}}{\log (1 - \frac{a \eta^r}{r^{r / 2}})} \bigg)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
\end{theorem}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $p$ and $q$.}
\KwResult{Community assignments $1, ..., K$.}
Compute $X$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
Apply single linkage clustering with $K$ communities on the ASE.\;
\caption{ASE clustering for nonintersecting communities.}
\end{algorithm}

@@

\begin{example}[Macaque visuotactile brain areas and connections]

\cite{https://doi.org/10.1111/j.1460-9568.2006.04678.x}

@@

>>>>

bibliography: bibliography.bib
output: rticles::asa_article
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
- \usepackage{amsfonts}
---

@@

\begin{definition}[Manifold block model]
Let $p, q \geq 0$, $d = p + q \geq 1$, $K \geq 1$, and $n \geq 1$ be integers.
If there are manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1] \}$, and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}$ such that each $x_i \in \mathcal{M}_k$ for some $k \in [K]$, then $A \sim \GRDPG_{p,q}(X; \rho_n)$ is a \emph{manifold block model}.
\end{definition}

In practice, we define each manifold $\mathcal{M}_k$ by a function $f_k : [0, 1]^r \to \mathcal{X}$, where $1 \geq r < d$ is the dimensionality of manifold $\mathcal{M}_k$.

@@

\begin{example}
Let $p = 2$, $q = 0$, $K = 2$, and $r = 1$. 
Define two manifolds by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \sim \Uniform(0, 1)$ and $z_1, ..., z_n \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $\begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \RDPG(X)$.

```{r fig1, fig.cap = 'Manifold block model described in Example 1. The latent configuration is on the left, and a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.', fig.height = 2.5, fig.width = 7}
set.seed(123456)
n1 <- 2 ** 5
n2 <- 2 ** 5
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))
u1 <- runif(n1)
u2 <- runif(n2)
x1 <- cos(pi / 3 * u1)
y1 <- sin(pi / 3 * u1)
x2 <- 1 + cos(pi / 3 * u2 + pi)
y2 <- 1 + sin(pi / 3 * u2 + pi)
data.matrix <- cbind(c(x1, x2), c(y1, y2))
A <- draw.graph(data.matrix %*% t(data.matrix))
Xhat <- embedding(A, 2, 0)

par(mfrow = c(1, 3))
# par(fig = c(0, .7, 0, 1))
plot(data.matrix, col = z * 2, asp = 1,
     xlab = NA, ylab = NA, xaxt = 'n', yaxt = 'n',
     xlim = c(-.1, 1.1), ylim = c(0, 1))
# ggplot() +
#   geom_point(aes(x = data.matrix[, 1], y = data.matrix[, 2], colour = factor(z))) +
#   labs(x = NULL, y = NULL) +
#   scale_colour_brewer(palette = 'Set1') +
#   theme(legend.position = 'none') +
#   coord_fixed()
# par(fig = c(.3, 1, 0, 1))
qgraph::qgraph(A, groups = factor(z), legend = FALSE)
plot(Xhat, col = z * 2, asp = 1, 
     xlab = NA, ylab = NA, xaxt = 'n', yaxt = 'n')
```
\end{example}

@@

Here, we provide two algorithms for MBM community detection. 
First, we consider the case in which communities correspond to manifolds in the latent space that do not intersect and are separated by some finite distance. In this scenario, we use the convergence of the ASE to show that single linkage clustering on the latent space produces a clustering such that the total number of misclustered vertices goes to zero, with high probability.

@@

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds without noise]
\label{nonintersect-no-noise}
Let $x_1, ..., x_n$ be points sampled on $K$ one-dimensional manifolds $\mathcal{M}_1, ..., \mathcal{M}_K$, defined by arc-length parameterizations $g_1, ..., g_K$. 
Suppose $\delta = \min_{k, \ell} \min_{t, s} \| g_k(t) - g_\ell(s) \| > 0$. 
Let $A_n$ be the event such that $\max_k \max_i \|x_{(i+1)}^{(k)} - x_{(i)}^{(k)}\| > \delta$, where $x_{(i)}^{(k)}$ is the $i^{th}$ order statistic on the $k^{th}$ manifold. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(A_n) < \epsilon$. 
\end{theorem}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds with noise]
\label{nonintersect-with-noise}
Suppose the same setup as in Theorem \ref{nonintersect-no-noise}. 
Define $y_i = x_i + e_i$ and $B_n$ as the event such that $\|y_{(i+1)}^{(k)} - y_{(i)}^{(k)}\| > \|y_{(i)}^{(k)} - y_{(j)}^{(\ell)}\|$ for some $i$, $j$, $k$, and $\ell$. 
Then if $\max_i \|e_i\| < \delta / 3$, for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(B_n) < \epsilon$. 
\end{theorem}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifold block models]
\label{nonintersect-mbm}
Suppose we have the same setup as in Theorem \ref{nonintersect-no-noise}. 
Let $p, q \geq 0$ such that $p + q = d$ where $d$ is the dimension of the range of each $g_k$. 
Suppose we draw $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some sparsity parameter $\rho_n \in (0, 1]$, and let $\hat{X}$ be the ASE of $A$. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, single linkage clustering produces zero community detection error with probability 1. 
\end{theorem}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $p$ and $q$.}
\KwResult{Community assignments $z_1, ..., z_n \in \{1, ..., K\}$.}
Compute $\hat{X}$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
Apply single linkage clustering with $K$ communities on $\hat{X}$\;
\caption{ASE clustering for nonintersecting communities.}
\end{algorithm}

@@

\begin{example}[Macaque visuotactile brain areas and connections \citep{https://doi.org/10.1111/j.1460-9568.2006.04678.x}]

@@

\appendix

\section{Proofs of Theorems}

To prove theorems \ref{nonintersect-no-noise}, \ref{nonintersect-with-noise}, and \ref{nonintersect-mbm}, we first need to introduce some theory for order statistics. 
Let $D_i = X_{(i+1)} - X_{(i)}$. Then if $\max_i D_i < \delta$, we have sufficient separation of points in $\mathcal{M}_1$. 
Then it is sufficient to quantify $P(\max_i D_i > \delta)$ as a function of $n$ and $\delta$ and show that this converges to zero as $n$ grows to $\infty$.

We now focus on the case where points are sampled uniformly on the unit interval to build up to more general distributions.

\begin{lemma}
Let $X_1, ..., X_n \iid \Uniform(0, 1)$. 
Then for any $\epsilon$ and $\delta > 0$, 
there exists an $N = O \big(\frac{-\log \epsilon}{\delta} \big)$ such that 
$P(\max_i X_{(i+1)} - X_{(i)} < \delta) \geq 1 - \epsilon$ when $n > N$.
\end{lemma}

Now we extend this lemma to general distributions on the unit interval.

\begin{lemma}
Let $X_1, ..., X_n \iid F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
Let $D_i = X_{(i+1)} - X_{(i)}$. 
Then for any $\epsilon > 0$, there exists $N > 0$ such that $P(\max_i D_i < \delta) \geq 1 - \epsilon$ when $n > N$.
\end{lemma}

This converges to $1$ in the limit $n \to \infty$.
\end{proof}

\begin{lemma}
\label{thm:multidim}
Let $X_1, ..., X_n \iid F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support. 
Define $E_n$ as the event that an $\eta$-neighborhood graph constructed from the sample is connected. 
Then for any $\epsilon > 0$, there exists $N = O \bigg( \frac{\log \epsilon \eta^r / r^{r/2}}{\log (1 - \frac{a \eta^r}{r^{r / 2}})} \bigg)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
\end{lemma}

### 2022-04-20 draft/draft.tex
still working on manuscript

\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\setcitestyle{numbers,square,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\usepackage{amsfonts}

@@

@@

@@

\hypertarget{preliminary-theory}{%
\subsubsection{Preliminary Theory}\label{preliminary-theory}}

\hypertarget{distributions-of-differences-of-order-statistics}{%
\paragraph{Distributions of differences of order
statistics}\label{distributions-of-differences-of-order-statistics}}

Let \(D_i = X_{(i+1)} - X_{(i)}\). Then if \(\max_i D_i < \delta\), we
have sufficient separation of points in \(\mathcal{M}_1\). Then it is
sufficient to quantify \(P(\max_i D_i > \delta)\) as a function of \(n\)
and \(\delta\) and show that this converges to zero as \(n\) grows to
\(\infty\).

\hypertarget{uniform-case}{%
\paragraph{Uniform case}\label{uniform-case}}

\begin{theorem}
Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Uniform}(0, 1)$. 
Then for any $\epsilon$ and $\delta > 0$, 
there exists an $N = O \big(\frac{-\log \epsilon}{\delta} \big)$ such that 
$P(\max_i X_{(i+1)} - X_{(i)} < \delta) \geq 1 - \epsilon$ when $n > N$.
\end{theorem}

\hypertarget{general-case-for-one-dimensional-manifolds}{%
\paragraph{General case for one-dimensional
manifolds}\label{general-case-for-one-dimensional-manifolds}}

\begin{theorem}
\label{thm:generalized}
Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
Let $D_i = X_{(i+1)} - X_{(i)}$. 
Then for any $\epsilon > 0$, there exists $N > 0$ such that $P(\max_i D_i < \delta) \geq 1 - \epsilon$ when $n > N$.
\end{theorem}

This converges to $1$ in the limit $n \to \infty$.

We can also approximate $F(1 - \delta) \approx 1 - a \delta$, which yields 
$1 - (n - 1) (1 - a \delta)^n$. Setting this $\geq 1 - \epsilon$ ...
\end{proof}

\hypertarget{extension-to-multidimensional-manifolds}{%
\paragraph{Extension to multidimensional
manifolds}\label{extension-to-multidimensional-manifolds}}

\begin{theorem}
\label{thm:multidim}
Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support. 
Define $E_n$ as the event that an $\eta$-neighborhood graph constructed from the sample is connected. 
Then for any $\epsilon > 0$, there exists $N = O \bigg( \frac{\log \epsilon \eta^r / r^{r/2}}{\log (1 - \frac{a \eta^r}{r^{r / 2}})} \bigg)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
\end{theorem}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $p$ and $q$.}
\KwResult{Community assignments $1, ..., K$.}
Compute $X$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
Apply single linkage clustering with $K$ communities on the ASE.\;
\caption{ASE clustering for nonintersecting communities.}
\end{algorithm}

@@

>>>>

\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\usepackage{amsfonts}

@@

\begin{definition}[Manifold block model]
Let $p, q \geq 0$, $d = p + q \geq 1$, $K \geq 1$, and $n \geq 1$ be integers.
If there are manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1] \}$, and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}$ such that each $x_i \in \mathcal{M}_k$ for some $k \in [K]$, then $A \sim \mathrm{GRDPG}_{p,q}(X; \rho_n)$ is a \emph{manifold block model}.
\end{definition}

In practice, we define each manifold \(\mathcal{M}_k\) by a function
\(f_k : [0, 1]^r \to \mathcal{X}\), where \(1 \geq r < d\) is the
dimensionality of manifold \(\mathcal{M}_k\).

@@

\begin{example}
Let $p = 2$, $q = 0$, $K = 2$, and $r = 1$. 
Define two manifolds by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \sim \mathrm{Uniform}(0, 1)$ and $z_1, ..., z_n \mathrm{Multinomial}(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $\begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \mathrm{RDPG}(X)$.

{\centering \includegraphics{draft_files/figure-latex/fig1-1}

\caption{Manifold block model described in Example 1. The latent configuration is on the left, and a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.}\label{fig:fig1}
\end{figure}
\end{example}

@@

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds without noise]
\label{nonintersect-no-noise}
Let $x_1, ..., x_n$ be points sampled on $K$ one-dimensional manifolds $\mathcal{M}_1, ..., \mathcal{M}_K$, defined by arc-length parameterizations $g_1, ..., g_K$. 
Suppose $\delta = \min_{k, \ell} \min_{t, s} \| g_k(t) - g_\ell(s) \| > 0$. 
Let $A_n$ be the event such that $\max_k \max_i \|x_{(i+1)}^{(k)} - x_{(i)}^{(k)}\| > \delta$, where $x_{(i)}^{(k)}$ is the $i^{th}$ order statistic on the $k^{th}$ manifold. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(A_n) < \epsilon$. 
\end{theorem}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds with noise]
\label{nonintersect-with-noise}
Suppose the same setup as in Theorem \ref{nonintersect-no-noise}. 
Define $y_i = x_i + e_i$ and $B_n$ as the event such that $\|y_{(i+1)}^{(k)} - y_{(i)}^{(k)}\| > \|y_{(i)}^{(k)} - y_{(j)}^{(\ell)}\|$ for some $i$, $j$, $k$, and $\ell$. 
Then if $\max_i \|e_i\| < \delta / 3$, for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(B_n) < \epsilon$. 
\end{theorem}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifold block models]
\label{nonintersect-mbm}
Suppose we have the same setup as in Theorem \ref{nonintersect-no-noise}. 
Let $p, q \geq 0$ such that $p + q = d$ where $d$ is the dimension of the range of each $g_k$. 
Suppose we draw $A \sim \mathrm{GRDPG}_{p,q}(X; \rho_n)$ for some sparsity parameter $\rho_n \in (0, 1]$, and let $\hat{X}$ be the ASE of $A$. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, single linkage clustering produces zero community detection error with probability 1. 
\end{theorem}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $p$ and $q$.}
\KwResult{Community assignments $z_1, ..., z_n \in \{1, ..., K\}$.}
Compute $\hat{X}$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
Apply single linkage clustering with $K$ communities on $\hat{X}$\;
\caption{ASE clustering for nonintersecting communities.}
\end{algorithm}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\appendix

\section{Proofs of Theorems}

To prove theorems \ref{nonintersect-no-noise},
\ref{nonintersect-with-noise}, and \ref{nonintersect-mbm}, we first need
to introduce some theory for order statistics. Let
\(D_i = X_{(i+1)} - X_{(i)}\). Then if \(\max_i D_i < \delta\), we have
sufficient separation of points in \(\mathcal{M}_1\). Then it is
sufficient to quantify \(P(\max_i D_i > \delta)\) as a function of \(n\)
and \(\delta\) and show that this converges to zero as \(n\) grows to
\(\infty\).

We now focus on the case where points are sampled uniformly on the unit
interval to build up to more general distributions.

\begin{lemma}
Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Uniform}(0, 1)$. 
Then for any $\epsilon$ and $\delta > 0$, 
there exists an $N = O \big(\frac{-\log \epsilon}{\delta} \big)$ such that 
$P(\max_i X_{(i+1)} - X_{(i)} < \delta) \geq 1 - \epsilon$ when $n > N$.
\end{lemma}

Now we extend this lemma to general distributions on the unit interval.

\begin{lemma}
Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
Let $D_i = X_{(i+1)} - X_{(i)}$. 
Then for any $\epsilon > 0$, there exists $N > 0$ such that $P(\max_i D_i < \delta) \geq 1 - \epsilon$ when $n > N$.
\end{lemma}

This converges to $1$ in the limit $n \to \infty$.
\end{proof}

\begin{lemma}
\label{thm:multidim}
Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support. 
Define $E_n$ as the event that an $\eta$-neighborhood graph constructed from the sample is connected. 
Then for any $\epsilon > 0$, there exists $N = O \bigg( \frac{\log \epsilon \eta^r / r^{r/2}}{\log (1 - \frac{a \eta^r}{r^{r / 2}})} \bigg)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
\end{lemma}
